---
title: "Data Wrangling"
author: "Brandeis Data Services"
date: "2025-01-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Install and Load Tidyverse Packages

You will need to `install.packages()` just once to download libraries
onto your computer.  After that, use `library()` any time you want to access
the tools and functions in a package.

```{r}
# To install packages:
#install.packages("tidyverse")

library(tidyverse)
```

## Load our data from a file into R environment

```{r}
penguins <- read_csv("penguins.csv")

```

##Tidy data
1. Each column is a variable
2. Each row is an observation
3. Each cell has a value

Let's explore our data set
```{r}
glimpse(penguins)
```

##Selecting columns
```{r}
select(penguins,species, body_mass_g, sex)

select(penguins, species:body_mass_g)

select(penguins, -year, -island)

select(penguins, starts_with("bill"))

select(penguins, ends_with("mm"))
```

##Filtering rows

```{r}
filter (penguins, island=="Torgersen")
filter (penguins, year<=2008)
filter (penguins, island=="Torgersen", sex=="male")
filter(penguins, species == "Chinstrap" | species == "Gentoo")
filter(penguins, species %in% c("Chinstrap", "Gentoo"))
```
#Missing values
```{r}
filter(penguins, is.na(sex))

filter(penguins, !is.na(sex))
```


You can use *pipes*. Pipes let you take the output of one function and send it 
directly to the next, which is useful when you need to do many things to the 
same dataset. 

Pipes in R look like `%>%` Read the pipe like the word “then”. 


```{r}
# Using pipes
penguins_biscoe<-penguins %>%
    filter(island == "Biscoe") %>%
    select(species, body_mass_g, sex)
```



#Exercise:
```{r}
# Create a new object with the data subset to include all species except Adelie and retain the species column and the ones relating to their bill
penguins %>%
    filter(species != "Adelie") %>%
    select(species, bill_length_mm, bill_depth_mm)

```


# Mutate

Frequently you’ll want to create new columns based on the values in existing 
columns, for example to do unit conversions, or to find the ratio of values in 
two columns. For this we’ll use mutate().

We might be interested in the body mass of penguins in kg instead of g:
```{r}
penguins %>%
    mutate(body_mass_kg = body_mass_g / 1000)
```


    
# Split-apply-combine data analysis and the summarize() function
Many data analysis tasks can be approached using the split-apply-combine 
paradigm: split the data into groups, apply some analysis to each group, and 
then combine the results. `dplyr` makes this very easy through the use of the 
`group_by()` function.

### The summarize() function
`group_by()` is often used together with `summarize()`, which collapses each 
group into a single-row summary of that group. 

`group_by()` takes as arguments:
- the column names that contain the categorical variables for which you want to calculate the summary statistics. 

So to compute the average body mass by species:
```{r}
penguins %>%
  group_by(species) %>% 
  summarize(body_mass_g_mean = mean(body_mass_g, na.rm=TRUE))
```


You can also group by multiple columns:
```{r}
penguins %>%
    group_by(island, species) %>%
    summarize(flipper_length_mm_mean = mean(flipper_length_mm, na.rm = TRUE),
              flipper_length_mm_min = min(flipper_length_mm, na.rm = TRUE),
              flipper_length_mm_max = max(flipper_length_mm, na.rm = TRUE),
              flipper_length_mm_sd = sd(flipper_length_mm, na.rm = TRUE))
```


# Counting
When working with data, we often want to know the number of observations found 
for each factor or combination of factors. For this task, `dplyr` provides 
`count()`. 

If we wanted to count the number of penguins by species, we would do:
```{r}
penguins %>%
  count(species)
```

    
For convenience, `count()` provides the `sort` argument to get results in decreasing order:
```{r}
penguins %>%
  count(species, sort = TRUE)

```

We can add more than one variable:
```{r}
penguins %>%
  count(species, island, sex)

```

#Arrange the order of your rows
The default is to arrange in ascending order. You can sse the desc() function on the variable inside arrange() to arrange in descending order
```{r}
penguins %>%
  arrange(body_mass_g)

```

##Renaming columns
```{r}
penguins %>% 
  rename(bill_length = bill_length_mm)
```
#case_when
```{r}
penguins %>%
  mutate(body_type = case_when(
       body_mass_g < 3000 ~ "small",
       body_mass_g >= 3000 & body_mass_g < 4500 ~ "normal",
       body_mass_g >= 4500 ~ "large"))
```

##Tall/Long vs. Wide data

Tall (or long) data are considered “tidy”, in that they adhere to the three tidy-data principles.
Sometimes, it might be more useful to have data in a "wide" format, where columns represent different values or levels of a variable. 
We can reshape our data using two tidyr functions, pivot_wider() and pivot_longer()
```{r}
penguins %>%
  pivot_longer(contains("_"))
```
Why would we want that? Perhaps we want to plot all the variables in a single ggplot call
```{r}
penguins %>%
  pivot_longer(contains("_")) %>% 
  ggplot(aes(y = value, 
             x = species,
             fill = species)) +
  geom_boxplot() +
  facet_wrap(~name, scales = "free_y")
```
We can clean up and change names during pivots
```{r}
penguins_long <- penguins %>% 
  pivot_longer(contains("_"),
               names_to = c("part", "measure" , "unit"),
               names_sep = "_",
               values_drop_na = TRUE)
penguins_long
```
#Pivoting data wider
First, we create a simple long datafame
```{r}
penguins_long_simple <- penguins %>% 
  pivot_longer(contains("_"))
penguins_long_simple
```
If we want to make this wider we can try the following steps:
```{r}
penguins_long_simple %>% 
  pivot_wider(names_from = name, 
              values_from = value)
```
The output looks weird, we need to add the row numbers to get uniquely identified values
```{r}
penguins_long_simple <- penguins %>%  
  mutate(sample = row_number()) %>%  
  pivot_longer(contains("_"),
               names_to = c("part", "measure" , "unit"),
               names_sep = "_",
               values_drop_na = TRUE)
penguins_long_simple
```

#Now, we can pivot
```{r}
pivot_wider<-penguins_long_simple %>% 
  pivot_wider(names_from = c("part", "measure", "unit"),
                         values_from = value)
```

# Using dplyr to merge tables

Let's create two data sets
```{r}
data1<- data.frame(country=c("Germany","Australia", "Ecuador"),
region=c("Europe","Western Pacific", "Americas" ),
life_exp=c(81, 83, 75))

data2<-data.frame(country=c("Germany","Australia", "Iceland"),
region=c("Europe","Western Pacific", "Europe" ),
urban_pop=c(76, 86, 94))

data1
data2
```

#"Joining" joins

What's is both data1 and data2?
```{r}
inner_join(data1, data2, by="country")
```



What's in data1 and bring with it the data that matches in data2
```{r}
left_join(data1, data2, by="country")
```



What's in data2 and bring with it the data that matches in data1
```{r}
right_join(data1, data2, by="country")
```



I want everything!
```{r}
full_join(data1, data2, by="country")
```



#Filtering joins

To filter one table against the rows of another
I want the rows in data1 that have a match in data2
```{r}
semi_join(data1, data2, by="country")
```


I want the rows in data1 that do not have a match in data2
```{r}
anti_join(data1, data2, by="country")
```


#Combining cases
```{r}
data3<- data.frame(country=c("Germany","Ecuador"),
life_exp=c(81,  75))

data4<-data.frame(country=c("Germany","Australia" ),
life_exp=c(81, 83 ))
```

One table on top of the other
```{r}
bind_rows(data3, data4)
```


Rows that appear in both
```{r}
intersect(data3, data4)
```



Rows that appear in data3 but not data4

```{r}
setdiff(data3, data4)
```

Rows that appear in data3 or data4


```{r}
union(data3, data4)

```



# Exporting data
Now that you have learned how to use dplyr to extract information from or 
summarize your raw data, you may want to export these new data sets to share 
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, 
there is a `write_csv()` function that generates CSV files from dataframes.
```{r}
help(write_csv)


```

